---
title: "Testing Sandbox"
author:
  - name: "Socorro Dominguez Vidaña"
    correspondence: false
    email: dominguezvid@wisc.edu
    orcid_id: 0000-0002-7926-4935
institute:
  - uwiscgeog:
    name: "University of Wisconsin -- Madison: Department of Geography"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  html_document:
    code_folding: show
    fig_caption: yes
    keep_md: yes
    self_contained: yes
    theme: readable
    toc: yes
    toc_float: yes
    css: "text.css"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, echo=FALSE}
options(warn = -1)
pacman::p_load(neotoma2, dplyr, ggplot2, sf, geojsonsf, leaflet, flextable, readr, stringr, tidyr, patchwork)
```

# Introduction

This document/binder is intended to act as a sandbox for new neotoma2 functions for SISAL.

We can work from a locally deployed API or from the development branch as of now.

```{r sitename}
# Setting the working environment. When the data is in production, this step can be ommitted.
Sys.setenv(APIPOINT = "https://str99atgzk.us-east-2.awsapprunner.com/v2.0/")
```

## Searching one cave
```{r}
kesang <- tryCatch({
  neotoma2::get_sites(sitename = "%Kesang Cave%")
}, error = function(e) {
  readRDS("data/kesang.RDS")
})
plotLeaflet(kesang)
```

## Summary of Cave

```{r summary}
neotoma2::summary(kesang)
```

```{r filling_data}
kesang_dl <- tryCatch({
  kesang %>% 
    neotoma2::filter(datasettype=="speleothem") %>% 
    neotoma2::get_downloads()
}, error = function(e) {
  readRDS("data/kesang_dl.RDS")
})
kesang_dl
```

## Getting the Speleothem Data
### `get_speleothems()`
```{r speleothem_data}
kesang_sp <- tryCatch({
  neotoma2::get_speleothems(kesang_dl)
}, error = function(e) {
  readRDS("data/kesang_sp.RDS")
})
kesang_sp
```

### `speleothems` slot is in `collunits` slot.

```{r speleothem_location}
kesang_sp[[1]]@collunits[[2]]@speleothems[[1]]
```
### Displaying Speleothem MetaData `speleothems()`
```{r show_speleothems_metadata}
neotoma2::speleothems(kesang_sp)
```


### Displaying sample/speleothem data `speleothem_details()`
```{r speleothem_data_detail}
data <- tryCatch({
  neotoma2::speleothemdetails(kesang_sp)
}, error = function(e) {
  read.csv("data/kesang_data.csv")
})

head(data)
```

## Taxa in Kesang
```{r taxa}
head(taxa(kesang_sp))
```

### Ordering the data in wide format `toWide()`

```{r toWide}
values <- kesang_sp[[1]]@collunits[[1]] %>%
  samples() %>%
  toWide(ecologicalgroup = c("ITOP"),
         unit = c("per mille"),
         groupby = "age",
         elementtype = NA,
         operation = "sum") %>%
  arrange(age) %>% na.omit()

values %>% 
  head() %>% 
  flextable() %>% 
  fontsize(size = 8, part = "all") %>%
  autofit() 
```


```{r data}
# Make sure age is in descending order (younger to older)
values <- values %>% arrange(desc(age))
head(values)
```


```{r plotIsotopesCode, message=FALSE, warning=FALSE, out.width='90%'}
# d18O plot (top)
p1 <- ggplot(values, aes(x = age, y = δ18O)) +
  geom_line(color = "steelblue") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{18}*O*" ("*"\u2030"*")"), x = NULL)

# d13C plot (bottom)
p2 <- ggplot(values, aes(x = age, y = δ13C)) +
  geom_line(color = "darkgreen") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{13}*C* "("*"\u2030"*")"), x = "Age (years BP)")

# Combine the two plots and add a title
(p1 / p2) + plot_annotation(title = paste('Data from',kesang_sp[[1]]$sitename))
```
