---
title: "Tests with Gael_2022 and Gael_2015"
author:
  - name: "Socorro Dominguez Vidaña"
    correspondence: false
    email: dominguezvid@wisc.edu
    orcid_id: 0000-0002-7926-4935
institute:
  - uwiscgeog:
    name: "University of Wisconsin -- Madison: Department of Geography"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  html_document
---

```{r setup, echo=FALSE}
options(warn = -1)
pacman::p_load(neotoma2, dplyr, ggplot2, sf, geojsonsf, leaflet, flextable, readr, stringr, tidyr, patchwork)
```

# Introduction

This document/binder is intended to act as a sandbox for new neotoma2 functions for SISAL.

We can work from a locally deployed API or from the development branch as of now.

```{r sitename}
# Setting the working environment. When the data is in production, this step can be ommitted.
Sys.setenv(APIPOINT = "https://api-dev.neotomadb.org/v2.0/")
```

## Searching one cave
```{r}
vallina <- tryCatch({
  neotoma2::get_sites(sitename = "%La Vallina%") %>%
    get_datasets()
}, error = function(e) {
  message("Could not reach to the API")
})
plotLeaflet(vallina)
```

## Summary of Cave

Since we applied `get_sites()` and `get_datasets()`, we can see all the datasets' IDs that are associated with Vallina Cave.

We can see what kinds of datatypes exist in each of the collection units.

In the neotoma model, all collection units should be in a site. The collection units are in the `collunits` slot of the `sites` object.

```{r summary}
neotoma2::summary(vallina)
```
## Getting Speleothem Data

### `get_downloads()` goes first

We can retrieve the speleothem data using the `get_speleothems()` function. But if we want to have all the data - including all sample data, we have to execute `get_downloads()` first. 

`speleothems` cannot be linked directly to `sites` because of the way Neotoma works, so `speleothems` are contained in the `collunits` slot.

This is only because of how the Neotoma model works. `get_speleothems()` will only modify the slots regarding `speleothems` data. But since that means overwriting the `collection unit` slot, we need to have the data downloaded first. If we do `get_downloads()` after, the `collection unit` slot will be overwritten again, and the `speleothems` data will be lost.

In order to have a smoother download, we can first filter the datasets to only include `speleothem` datasets, and then download only those.
```{r filling_data}
vallina_dl <- vallina %>% 
    neotoma2::filter(datasettype=="speleothem") %>% 
    neotoma2::get_downloads()
vallina_dl
```

### `get_speleothems()` to retrieve `speleothem` metadata

```{r speleothem_data}
vallina_sp <- neotoma2::get_speleothems(vallina_dl)
vallina_sp
```
We only have one site, **Vallina Cave**. This site has multiple collection units, and each collection unit can have one or more speleothems associated with it.

To retrieve the speleothem metadata, we can access the `speleothems` slot within each `collunits` slot.

### `speleothems` slot is in `collunits` slot.

We can see all the `collection units` in our site by doing:

```{r speleothem_collunits}
vallina_sp[[1]]@collunits
```

Then, we can explore each speleothem by iterating over the `collunits` slot AND the `speleothems` slot. 
Most of the times, one `collunit` only has one `speleothem` - but there could be more, or we could add more with the `set_` functions.
```{r speleothem_metadata}
# One one speleothem
vallina_sp[[1]]@collunits[[1]]@speleothems[[1]]
```
```{r speleothem_metadata2}
# One one speleothem
all_gael <- vallina_sp[[1]]@collunits[[5]]@speleothems
all_gael
```

In this case, if we want to see data specific to only one of the speleothems, we have to access one by one:

```{r speleothem_metadata3}
# Retrieving only Gael_2022
gael2022 <- vallina_sp[[1]]@collunits[[5]]@speleothems[[1]]
gael2022
```

### Displaying Speleothem MetaData `speleothems()`

The function `speleothems()` aims to retrieve the speleothem metadata that exists in Neotoma.
It can be applied to only one or multiple collection units:

```{r `speleothems()`_on_a_collunit}
gael_cu <- vallina_sp[[1]]@collunits[[5]]
neotoma2::speleothems(gael_cu)
```
The problem here is, that `speleothems` does not link back to `datasets`. That is why it would seem like both entities point back to the same two dataset IDs. We need to create a path so that the `datasetID` can also be linked to it's own speleothem.

Dataset **68606** belongs to **Gael_2022** and Dataset **69328** belongs to **Gael_2015** (This information is from the upload logs).


```{r `speleothems()`_on_a_site}
neotoma2::speleothems(vallina_sp)
```


### Displaying sample/speleothem data `speleothem_details()`
```{r speleothem_data_detail}
data <- neotoma2::speleothemdetails(vallina_sp)
head(data)
```

Finding speleothems by Sr_Ca data:

```{r find_speleothem_by_SrCa}
get_sites(taxa='Strontium:Calcium', all_data=T) %>%
  get_datasets() %>%
  neotoma2::filter(datasettype == "speleothem") %>%
  neotoma2::summary()
```

## Taxa in Vallina - different function name, but same behaviour.
```{r renaming taxa}
type_of_data <- taxa
```

```{r type_of_data}
head(type_of_data(vallina_sp))
```

### Ordering the data in wide format `toWide()`

```{r toWide}
values <- vallina_sp[[1]]@collunits[[1]] %>%
  samples() %>%
  toWide(ecologicalgroup = c("ITOP"),
         unit = c("per mille"),
         groupby = "age",
         elementtype = NA,
         operation = "sum") %>%
  arrange(age) %>% na.omit()

values %>% 
  head() %>% 
  flextable() %>% 
  add_header_lines(values=vallina_sp[[1]]@collunits[[1]]@handle) %>%
  fontsize(size = 8, part = "all") %>%
  autofit() 
```


```{r data values}
# Make sure age is in descending order (younger to older)
values <- values %>% arrange(desc(age))
head(values)
```


```{r plotIsotopesCode, message=FALSE, warning=FALSE, out.width='90%'}
# d18O plot (top)
p1 <- ggplot(values, aes(x = age, y = δ18O)) +
  geom_line(color = "steelblue") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{18}*O*" ("*"\u2030"*")"), x = NULL)

# d13C plot (bottom)
p2 <- ggplot(values, aes(x = age, y = δ13C)) +
  geom_line(color = "darkgreen") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{13}*C* "("*"\u2030"*")"), x = "Age (years BP)")

# Combine the two plots and add a title
(p1 / p2) + plot_annotation(title = paste('Data from',vallina_sp[[1]]$collunits[[1]]@handle, "(", vallina_sp[[1]]$sitename, ")"))
```
------

## Red Flags regarding Gael_2015 and Gael_2022

Because the speleothem data is associated to the collection unit, I do not have a way to trace back to which datasetID belongs to each speleothemID. I think this is something that we can correct in the upload process by changing the speleothems-collectionunits table to be speleothems-datasets table (or add a new one). 

The data information exists for both datasets:

### Dataset **68606** belongs to **Gael_2022** entity **803**

```{r download_68606}
gael2022_dl <- neotoma2::get_downloads(68606)
gael2022_sp <- neotoma2::get_speleothems(gael2022_dl)
neotoma2::speleothems(gael2022_sp)
```
You can see that the `datasetID` is being associated to both `entityID`s.

Let's filter out for clarity:

```{r download_68606_filtering_803}
gael2022_dl <- neotoma2::get_downloads(68606)
gael2022_sp <- neotoma2::get_speleothems(gael2022_dl)
neotoma2::speleothems(gael2022_sp) %>% filter(entityid == 803)
```

```{r speleothem_68606_data_speleo}
speleothem_data_68606 <- neotoma2::speleothemdetails(gael2022_sp)
speleothem_data_68606 <- speleothem_data_68606 %>% dplyr::filter(entityid == 803) # To avoid duplicates since it is associated to two entities
speleothem_data_68606 %>% head()
```

```{r checks_68606}
samples_68606 <- neotoma2::samples(gael2022_sp)
samples_68606 %>% 
  filter(variablename == "δ13C") %>%
  summarise(d13min = all(value >= -8.16),
            d13max = all(value <= -3.83),
            depth = all(depth >= 779)) %>%
  mutate(entityname = speleothem_data_68606$entityname %>% unique()) %>%
  select(entityname, depth, d13min, d13max)
```

### Dataset **69328** belongs to **Gael_2015**  or entity **804**

```{r download_69328_filtering_804}
gael2015_dl <- neotoma2::get_downloads(69328)
gael2015_sp <- neotoma2::get_speleothems(gael2015_dl)
neotoma2::speleothems(gael2015_sp) %>% filter(entityid == 804)
```

```{r speleothem_69328_data}
speleothem_data_69328 <- neotoma2::speleothemdetails(gael2015_sp)
speleothem_data_69328 <- speleothem_data_69328 %>% dplyr::filter(entityid == 804) # To avoid duplicates since it is associated to two entities
speleothem_data_69328 %>% head()
```

```{r checks_69328}
samples_69328 <- neotoma2::samples(gael2015_sp)
samples_69328 %>% 
  filter(variablename == "δ13C") %>%
  summarise(d13min = all(value >= -9.72),
            d13max = all(value <= -3.84),
            depth = all(depth <= 841)) %>%
  mutate(entityname = speleothem_data_69328$entityname %>% unique()) %>%
  select(entityname, depth, d13min, d13max)
```
**Conclusion regarding the red flag:** From the Database perspective, the insertions into Neotoma are correct. However, the problem lies that since the `speleothems` can only be traced back to their `collection unit`, and not to their `datasetID`, we cannot separate the data properly.

The duplication happens because if a collectionunit has more than 1 speleothem entities, then, the collectionunit can have the dataset more times, one for each speleothem entity.
