---
title: "Tests with Gael_2022 and Gael_2015"
author:
  - name: "Socorro Dominguez Vidaña"
    correspondence: false
    email: dominguezvid@wisc.edu
    orcid_id: 0000-0002-7926-4935
institute:
  - uwiscgeog:
    name: "University of Wisconsin -- Madison: Department of Geography"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  html_document
---

```{r setup, echo=FALSE}
options(warn = -1)
pacman::p_load(neotoma2, dplyr, ggplot2, sf, geojsonsf, leaflet, flextable, readr, stringr, tidyr, patchwork)
```

# Introduction

This document/binder is intended to act as a sandbox for new neotoma2 functions for SISAL.

We can work from a locally deployed API or from the development branch as of now.

```{r sitename}
# Setting the working environment. When the data is in production, this step can be ommitted.
Sys.setenv(APIPOINT = "https://api-dev.neotomadb.org/v2.0/")
```
## Sites with speleothem data:
```{r}
speleost <- get_datasets(datasettype = "%speleothem%")
speleost
```

## Searching one cave
```{r}
vallina <- get_datasets(sitename = "%La Vallina%")
plotLeaflet(vallina)
```

## Summary of Cave

Since we applied `get_sites()` and `get_datasets()`, we can see all the datasets' IDs that are associated with Vallina Cave.

We can see what kinds of datatypes exist in each of the collection units.

In the neotoma model, all collection units should be in a site. The collection units are in the `collunits` slot of the `sites` object.

```{r summary}
neotoma2::summary(vallina)
```
## Getting Speleothem Data

### `get_downloads()` goes first

We can retrieve the speleothem data using the `get_speleothems()` function. But if we want to have all the data - including all sample data, we have to execute `get_downloads()` first. 

`speleothems` cannot be linked directly to `sites` because of the way Neotoma works, so `speleothems` are contained in the `collunits` slot.

This is only because of how the Neotoma model works. `get_speleothems()` will only modify the slots regarding `speleothems` data. But since that means overwriting the `collection unit` slot, we need to have the data downloaded first. If we do `get_downloads()` after, the `collection unit` slot will be overwritten again, and the `speleothems` data will be lost.

In order to have a smoother download, we can first filter the datasets to only include `speleothem` datasets, and then download only those.
```{r filling_data}
vallina_dl <- vallina %>% 
    neotoma2::filter(datasettype=="speleothem") %>% 
    neotoma2::get_downloads()
vallina_dl
```

### `get_speleothems()` to retrieve `speleothem` metadata

```{r speleothem_data}
vallina_sp <- neotoma2::get_speleothems(vallina_dl)
vallina_sp
```
We only have one site, **Vallina Cave**. This site has multiple collection units, and each collection unit can have one or more speleothems associated with it.

To retrieve the speleothem metadata, we can access the `speleothems` slot within each `collunits` slot.

### `speleothems` slot is in `collunits` slot.

We can see all the `collection units` in our site by doing:

```{r speleothem_collunits}
vallina_sp[[1]]@collunits
```

Then, we can explore each speleothem by iterating over the `collunits` slot AND the `speleothems` slot. 
Most of the times, one `speleothem` only has one `collunit` but it could be the case that it has more than one collunit. However, a `collection unit` can only belong to one `speleothem.`
```{r speleothem_metadata}
# One one speleothem
vallina_sp[[1]]@collunits[[1]]@speleothems[[1]]
```
```{r speleothem_metadata2}
# One one speleothem
vallina_sp[[1]]@collunits[[2]]@speleothems[[1]]
```

In this case, if we want to see data specific to only one of the speleothems, we have to access one by one:

```{r speleothem_metadata3}
# Retrieving only Gael_2015
gael2015 <- vallina_sp[[1]]@collunits[[1]]@speleothems[[1]]
gael2015
```

### Displaying Speleothem MetaData `speleothems()`

The function `speleothems()` aims to retrieve the speleothem metadata that exists in Neotoma.
It can be applied to only one or multiple collection units:

```{r `speleothems()`_on_a_collunit}
gael_cu <- vallina_sp[[1]]@collunits[[1]]
neotoma2::speleothems(gael_cu)
```
Applying to the whole site:

```{r `speleothems()`_on_a_site}
neotoma2::speleothems(vallina_sp)
```


### Displaying sample/speleothem data `speleothem_details()`
```{r speleothem_data_detail}
data <- neotoma2::speleothemdetails(vallina_sp)
head(data)
```

Finding speleothems by Sr_Ca data:

```{r find_speleothem_by_SrCa}
get_sites(taxa='Strontium:Calcium', all_data=T) %>%
  get_datasets() %>%
  neotoma2::filter(datasettype == "speleothem") %>%
  neotoma2::summary()
```

## Taxa in Vallina - different function name, but same behaviour.
```{r renaming taxa}
type_of_data <- taxa
```

```{r type_of_data}
head(type_of_data(vallina_sp))
```

### Ordering the data in wide format `toWide()`

```{r toWide}
values <- vallina_sp[[1]]@collunits[[1]] %>%
  samples() %>%
  toWide(ecologicalgroup = c("ITOP"),
         unit = c("per mille"),
         groupby = "age",
         elementtype = NA,
         operation = "sum") %>%
  arrange(age) %>% na.omit()

values %>% 
  head() %>% 
  flextable() %>% 
  add_header_lines(values=vallina_sp[[1]]@collunits[[1]]@handle) %>%
  fontsize(size = 8, part = "all") %>%
  autofit() 
```


```{r data values}
# Make sure age is in descending order (younger to older)
values <- values %>% arrange(desc(age))
head(values)
```


```{r plotIsotopesCode, message=FALSE, warning=FALSE, out.width='90%'}
# d18O plot (top)
p1 <- ggplot(values, aes(x = age, y = δ18O)) +
  geom_line(color = "steelblue") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{18}*O*" ("*"\u2030"*")"), x = NULL)

# d13C plot (bottom)
p2 <- ggplot(values, aes(x = age, y = δ13C)) +
  geom_line(color = "darkgreen") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{13}*C* "("*"\u2030"*")"), x = "Age (years BP)")

# Combine the two plots and add a title
(p1 / p2) + plot_annotation(title = paste('Data from',vallina_sp[[1]]$collunits[[1]]@handle, "(", vallina_sp[[1]]$sitename, ")"))
```
------

## Red Flags regarding Gael_2015 and Gael_2022 **Fixed**

Because the speleothem data is associated to the collection unit, I was wrongly associating the speleothem to the collection unit instead of associating the collection unit to the speleothem. I have then fixed this. However, I cannot use `persist_id` as handle as the handle must be UNIQUE.

### Dataset **66768** belongs to **Gael_2015**  or entity **804**

```{r download_66768_filtering_804}
gael2015_dl <- neotoma2::get_downloads(66768)
gael2015_sp <- neotoma2::get_speleothems(gael2015_dl)
neotoma2::speleothems(gael2015_sp)
```

```{r speleothem_66768_data}
speleothem_data_66768 <- neotoma2::speleothemdetails(gael2015_sp)
speleothem_data_66768 %>% head()
```

```{r checks_66768}
samples_66768 <- neotoma2::samples(gael2015_sp) %>%
  left_join(speleothem_data_66768 %>% 
              select(entityid, entityname, datasetid, sampleid),
            by = c("datasetid", "sampleid"))
samples_66768 %>% 
  filter(variablename == "δ13C") %>%
  summarise(d13min = all(value >= -9.7225),
            d13max = all(value <= -3.8325),
            depth = all(depth <= 841),
            entityid = all(entityid == 804),
            datasetid = all(datasetid == 66768)) %>%
  mutate(entityname = samples_66768$entityname %>% unique()) %>%
  select(entityname, depth, d13min, d13max, entityid, datasetid)
```


### Dataset **66770** belongs to **Gael_2022** entity **803**

```{r download_66770}
gael2022_dl <- neotoma2::get_downloads(66770)
gael2022_sp <- neotoma2::get_speleothems(gael2022_dl)
neotoma2::speleothems(gael2022_sp)
```
You can see that the `datasetID` is being associated to both `entityID`s.


```{r speleothem_66770_data_speleo}
speleothem_data_66770 <- neotoma2::speleothemdetails(gael2022_sp)
speleothem_data_66770 %>% head()
```

```{r}
chronologies(gael2022_sp)
```

```{r}
samples(gael2022_sp) %>% head()
```


```{r checks_66770}
samples_66770 <- neotoma2::samples(gael2022_sp) %>%
  left_join(speleothem_data_66770 %>% 
              select(entityid, entityname, datasetid, sampleid),
            by = c("datasetid", "sampleid"))

samples_66770 %>% 
  filter(variablename == "δ13C") %>%
  summarise(d13min = all(value >= -8.16),
            d13max = all(value <= -3.83),
            depth = all(depth >= 779),
            entityid = all(entityid == 803),
            datasetid = all(datasetid == 66770)) %>%
  mutate(entityname = samples_66768$entityname %>% unique()) %>%
  select(entityname, depth, d13min, d13max, entityid, datasetid)
```
### Downloading Gael_2015 and Gael_2022 at the same time

```{r dl_all_gael}
vallina <- get_sites(sitename = "%La Vallina%") %>%
  get_datasets()

vallina_dl <- vallina %>% 
    neotoma2::filter(datasettype == "speleothem") %>% 
    neotoma2::get_downloads()
```

```{r adding_speleodata}
vallina_sp <- neotoma2::get_speleothems(vallina_dl)
neotoma2::speleothems(vallina_sp)
```

```{r speleothemdetails}
neotoma2::speleothemdetails(vallina_sp) %>% head()
```
### Checks for new downloaded object:
```{r all_details}
vallina_speleo_details <- speleothemdetails(vallina_sp)
vallina_samples <- samples(vallina_sp)

testing <- vallina_samples %>%
  left_join(vallina_speleo_details %>% 
              select(entityid, entityname, datasetid, sampleid),
            by = c("datasetid", "sampleid"))

testing803 <- testing %>% 
  filter(entityid == 803,
         variablename == "δ13C") %>%
  summarise(d13min = all(value >= -8.16),
            d13max = all(value <= -3.83),
            depth = all(depth >= 779),
            datasetid = all(datasetid == 66770),
            entityid = all(entityid == 803)) %>%
  mutate(entityid_c = "803") %>%
  select(entityid_c, depth, d13min, d13max, datasetid, entityid)
testing803
```
```{r testing804}
testing804 <- testing %>% 
  filter(entityid == 804,
         variablename == "δ13C") %>%
  summarise(d13min = all(value >= -9.7225),
            d13max = all(value <= -3.8325),
            depth = all(depth <= 841),
            entityid = all(entityid == 804),
            datasetid = all(datasetid == 66768)) %>%
  mutate(entityid_c = "804") %>%
  select(entityid_c, depth, d13min, d13max, datasetid, entityid)
testing804
```


**Conclusion regarding the red flag:** I have corrected the database so that one speleothem can have multiple collection units but one collection unit can only have one speleothem.

### Different sample ages for all chronologies:

```{r chronologies_all}
vallina_sp[[1]]@collunits[[1]]@datasets[[1]]@samples[[1]]@ages
```
To see the full chronologies of the first site of `vallina_sp`:
```{r full_chronologies}
head(as.data.frame(chronologies(vallina_sp[[1]])))
```
I can extract all of the chronology IDs and filter to only retrieve the Radiocarbon dates:

```{r all_chronologies}
# Retrieve only chronologies IDs for Radiocarbon dates
radiocarbon_ids <- as.data.frame(chronologies(vallina_sp)) %>%
  distinct() %>%
  pull(chronologyid)
```


Retrieve ALL the samples ages:
```{r all_sample_ages}
all_sample_ages <- purrr::map(vallina_sp@sites, function(vallina_sp) {
  purrr::map(vallina_sp@collunits, function(cu) {
    purrr::map(cu@datasets, function(ds) {
      purrr::map(ds@samples, function(samp) {
        df <- samp@ages[samp@ages$chronologyid %in% radiocarbon_ids, ]
        if (nrow(df) != 0) {
          df$sampleid <- samp@sampleid
          #force agetype as character
          df <- df %>% mutate(agetype = as.character(agetype),
                              dsid = ds@datasetid,
                              cuid = cu@collectionunitid,
                              siteid = vallina_sp@siteid)
          return(df)
        }
      }) %>%
        bind_rows()
    }) %>%
      bind_rows()
  }) %>%
    bind_rows()
}) %>% 
  bind_rows() %>%
  select(chronologyname, siteid, cuid, dsid, sampleid, 
         age, agetype, ageolder, ageyounger, 
         chronologyid)
head(all_sample_ages)
```

---

## Other Available Speleothems

```{r}
speleost <- get_datasets(datasettype = "%speleothem%")
speleost <- get_speleothems(speleost)
head(speleost)
```

```{r}
head(speleothems(speleost))
```

```{r}
head(speleothemdetails(speleost))
```