---
title: "Testing Sandbox"
author:
  - name: "Socorro Dominguez Vidaña"
    correspondence: false
    email: dominguezvid@wisc.edu
    orcid_id: 0000-0002-7926-4935
institute:
  - uwiscgeog:
    name: "University of Wisconsin -- Madison: Department of Geography"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  html_document:
    code_folding: show
    fig_caption: yes
    keep_md: yes
    self_contained: yes
    theme: readable
    toc: yes
    toc_float: yes
    css: "text.css"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, echo=FALSE}
options(warn = -1)
pacman::p_load(neotoma2, dplyr, ggplot2, sf, geojsonsf, leaflet, flextable, readr, stringr, tidyr, patchwork)
```

# Introduction

This document/binder is intended to act as a sandbox for new neotoma2 functions for SISAL.

We can work from a locally deployed API or from the development branch as of now.

```{r sitename}
# Setting the working environment. When the data is in production, this step can be ommitted.
Sys.setenv(APIPOINT = "https://api-dev.neotomadb.org/v2.0/")
```

## Searching one cave
```{r}
kesang <- tryCatch({
  neotoma2::get_sites(sitename = "%Kesang Cave%")
}, error = function(e) {
  NULL
})
plotLeaflet(kesang)
```

## Summary of Cave

```{r summary}
neotoma2::getids(kesang)
```

```{r filling_data}
kesang_dl <- tryCatch({
  kesang %>% 
    neotoma2::filter(datasettype=="speleothem") %>% 
    neotoma2::get_downloads()
}, error = function(e) {
  readRDS("data/kesang_dl.RDS")
})
kesang_dl
```

## Getting the Speleothem Data
### `get_speleothems()`
```{r speleothem_data}
kesang_sp <- tryCatch({
  neotoma2::get_speleothems(kesang_dl)
}, error = function(e) {
  readRDS("data/kesang_sp.RDS")
})
kesang_sp
```

### `speleothems` slot is in `collunits` slot.

```{r speleothem_location}
kesang_sp[[1]]@collunits[[2]]@speleothems[[1]]
```
### Displaying Speleothem MetaData `speleothems()`
```{r show_speleothems_metadata}
neotoma2::speleothems(kesang_sp)
```


### Displaying sample/speleothem data `speleothem_details()`
```{r speleothem_data_detail}
data <- tryCatch({
  neotoma2::speleothemdetails(kesang_sp)
}, error = function(e) {
  read.csv("data/kesang_data.csv")
})

head(data)
```

## Taxa in Kesang
```{r taxa}
head(taxa(kesang_sp))
```

### Ordering the data in wide format `toWide()`

```{r toWide}
values <- kesang_sp[[1]]@collunits[[1]] %>%
  samples() %>%
  toWide(ecologicalgroup = c("ITOP"),
         unit = c("per mille"),
         groupby = "age",
         elementtype = NA,
         operation = "sum") %>%
  arrange(age) %>% na.omit()

values %>% 
  head() %>% 
  flextable() %>% 
  fontsize(size = 8, part = "all") %>%
  autofit() 
```


```{r data}
# Make sure age is in descending order (younger to older)
values <- values %>% arrange(desc(age))
head(values)
```


```{r plotIsotopesCode, message=FALSE, warning=FALSE, out.width='90%'}
# d18O plot (top)
p1 <- ggplot(values, aes(x = age, y = δ18O)) +
  geom_line(color = "steelblue") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{18}*O*" ("*"\u2030"*")"), x = NULL)

# d13C plot (bottom)
p2 <- ggplot(values, aes(x = age, y = δ13C)) +
  geom_line(color = "darkgreen") +
  #scale_x_reverse() +
  theme_minimal() +
  labs(y = expression(delta^{13}*C* "("*"\u2030"*")"), x = "Age (years BP)")

# Combine the two plots and add a title
(p1 / p2) + plot_annotation(title = paste('Data from',kesang_sp[[1]]$sitename))
```

------

I have copied some of the exercises in the simple workflow using the new neotoma2 functions for speleothems. 

Please review them and let me know if there is any issue.

Following some functions of the simple_workflow:

```{r boundingBox}
geoJSON <- '{"type": "Polygon",
        "coordinates": [[
            [-7.030,  36.011],
            [-18.807, 23.537],
            [-19.247, 10.282],
            [-9.139,  -0.211],
            [18.370, -37.546],
            [35.069, -36.352],
            [49.571, -27.097],
            [58.185,   0.755],
            [53.351,  13.807],
            [43.946,  12.008],
            [31.202,  33.629],
            [18.897,  34.648],
            [12.393,  35.583],
            [11.075,  38.184],
            [-7.030,  36.011]
          ]
        ]}'

africa_sf <- geojsonsf::geojson_sf(geoJSON)

# Note here we use the `all_data` flag to capture all the sites within the polygon.
# We're using `all_data` here because we know that the site information is relatively small
# for Africa. If we were working in a new area or with a new search we would limit the
# search size.
africa_sites <- neotoma2::get_sites(loc = africa_sf, all_data = FALSE)
africa_sites
```


```{r datasetsFromSitesResult, message=FALSE}
africa_datasets <- neotoma2::get_datasets(loc = africa_sf, all_data = TRUE)

datasets(africa_datasets) %>% 
  as.data.frame() %>% 
  head() %>% 
  flextable() %>% 
  fontsize(size = 8, part = "all") %>%
  width(width = 1.2) %>%
  autofit() 
```


```{r downloadsCode}
africa_speleothem <- africa_datasets %>% 
  neotoma2::filter(datasettype == 'speleothem')

neotoma2::summary(africa_speleothem) %>% 
  head() %>% 
  flextable() %>% 
  fontsize(size = 8, part = "all") %>%
  autofit() 
```

```{r taxaCode}
africa_dl <- africa_speleothem %>% get_downloads() 

as.data.frame(africa_dl) %>% 
  head() %>% 
  flextable() %>% 
  fontsize(size = 8, part = "all") %>%
  width(width = 1.2) %>%
  autofit() 
```

```{r add_entityCode}
africa_sp <- get_speleothems(africa_dl)
africa_sp
```

```{r samp_w_entityCode}
samp_w_entity <- speleothems(africa_sp)
samp_w_entity
```



```{r taxa2Code, message=FALSE}
neotomatx <- africa_dl %>% taxa()
neotomatx %>% 
  head() %>% 
  flextable() %>% 
  fontsize(size = 8, part = "all")  %>%
  width(width = 1.2) %>%
  autofit() 
```